A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-08-08 sections/02-apis.tex
Included missing reference

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin
a parallel region.
Listing \ref{lst:listing-pthreads} presents a sample code using the
\textit{Pthreads} library. The program begins execution in a
single thread and until line 6 only one thread is running.

When \texttt{pthread\_create} is called a thread is created and the application
enters a parallel region with two simultaneous threads.  The main mechanism
available for synchronizing thread executions is the barrier. The
\textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 11 shows an example of thread synchronization using a barrier.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:listing-pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int main(){
    pthread_t thread_id;
    DATATYPE *parameters = 10;
    int return_code;

    return_code = pthread_create(&thread_id, NULL,
                                 task_function,
                                 (void *) parameters);

    printf("Return Code: %d\n", return_code );
    pthread_join(thread_id , NULL);

    return 0;
}

void *task_function(void *data){
    DATATYPE *value;
    value = (DATATYPE *) data;
}
\end{lstlisting}

>>>>

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin
a parallel region.
Listing \ref{lst:listing-pthreads} presents a sample code using the
\textit{Pthreads} library. The code was taken from
\cite{goncalves:OpenMPNotEasy} and shows a simple multiplication of two
matrices element-wise and summation of this multiplication.

The program begins its execution in a single thread and until line 17 only one
thread is running. When \texttt{pthread\_create} is called a thread is created
and the application enters a parallel region with two simultaneous threads. As
it happens inside a for loop, several threads are expected to be created.

The strategy used for such parallelization is to devide the vectors in chunks
and delegating the multiplication of the elements inside these chunks among
threads, what can be seen in lines 14 and 15.

The main mechanism available for synchronizing thread executions is the
barrier. The \textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 21 shows an example of thread synchronization using a barrier.

As the threads are supposed to write in a shared variable (\texttt{dot}), we
have to explicitly take care of the concurrent access to the variable, as seen
in Line 33 and 35, where the threads have to perform a lock at a mutex in order
to enter the critical region.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:listing-pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int A[SIZE], B[SIZE]; int dot = 0;
pthread_mutex_t lock ;
int main(int argc, char *argv []){ 
    int num_threads, i, part size ;
    pthread_t tid [MAX_THREADS];
    pthread_attr_t attr; 
    pthread_attr_init(&attr);
    /* Initialization vectors was supressed . */
    part_size = (SIZE + num_threads - 1) / num_threads;
    for (i=0; i < num_threads; i++){
        struct arg_struct *args = 
                malloc(sizeof(struct arg_struct));
        args->id = i;
        args->i_part = (i * part_size);
        args->e_part = MIN(((i + 1)* (part_size)), 
                           SIZE) ;
        pthread_create(&(tid[i]) ,&attr, vecDot, 
                       (void *) args) ;
    }
    for (i=0; i<num_threads; i++) { 
        pthread_join(tid[i], NULL);
    }
    printf("Final result: %d.\n", dot); 
    return 0;
}
void *vecDot (void *arguments){
    struct arg_struct *args = arguments;
    int i, partial_dot=0;
    for(i = args->i_part; i < args->e_part; i++){
        partial_dot += A[i] * B[i]; 
    }
    /* Critical region . */ 
    pthread_mutex_lock(&lock) ; 
    dot += partial_dot;
    pthread_mutex_unlock(&lock) ;
    pthread_exit(0); 
}
\end{lstlisting}
